Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-07 paint-contour-tree.tex
Various small changes

\begin{definition}
 Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
 is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
 For $p\in P(T)$, we use $r_p$ to denote the lowest vertex in $p$ (i.e. its root) and 
 $|p|$ to denote its length.

@@

\begin{definition}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
 For $p\in P(T)$, we say that $p$ is \emph{light} if $4 |p| < |H_{r_p}|$, and we say it is \emph{heavy} otherwise.
\end{definition}

@@

\begin{definition}
 We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
\end{definition}

@@

\begin{lemma}
\label{lem:adjacent}
 Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.

 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
 Then $\sum_{i=1}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
 %Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
\end{lemma}
\begin{proof}
 %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
 %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
 %Observe that $\sum_{i=0}^k d_i = l_p$.
 %
 % 
 %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=1}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=1}^k |H_i'|\leq |H_p|$.
 Putting these together gives
 \[
  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
 \]
\end{proof}

\begin{corollary}
 Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
 Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
\end{corollary}
\begin{proof}
 Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
 Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$.  Call a light path in $P(T)$ reachable from $p$ if there is a path in the tree
 $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
 Call this set of reachable paths the \emph{light path tree} of $p$.  Naturally this set can be viewed as a tree 
 (i.e. $p$ is the root of this tree and each other path has a well defined depth from $p$).
\end{definition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Junk, don't read}
The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.

>>>>

\begin{definition}
 Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
 is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
 For $p\in P(T)$, we use $|p|$ to denote its length and $r_p$ to denote the lowest vertex in $p$, which we call the \emph{root} of $p$.

@@

\begin{definition}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
 For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
\end{definition}

@@

\begin{definition}
Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$, corresponding to 
the path decomposition $P(T)$.

We say two path in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree. 
\end{definition}

%\begin{definition}
%We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
%Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
%$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
%Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
%\end{definition}

@@

\begin{lemma}
\label{lem:adjacent}
 Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
 Also, let $H_0$ denote the heap at the leaf of $p$.

 Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
\end{lemma}
\begin{proof}
 %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
 %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
 %Observe that $\sum_{i=0}^k d_i = l_p$.
 %
 % 
 %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
 Putting these together gives
 \[
  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
 \]
\end{proof}

Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.

THERE IS A PROBLEM in the following lemma.

\begin{lemma}
Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
$\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
\end{lemma}
\begin{proof}
By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
 As $x\log x$ is a convex function, Jensen's inequality then implies 
 $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
\end{proof}

\begin{corollary}
\label{cor:heavyLight}
 Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
 Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
\end{corollary}
\begin{proof}
 Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
 Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$.  Call a light path in $P(T)$ reachable from $p$ if there is a path in the tree
 $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
 Call this set of reachable paths the \emph{light path set} of $p$.  
 Consider the natural tree over the paths in the light path set of $p$, where each vertex in the tree is a light path 
 $p$ is the root.  We call this tree the \emph{light path tree} of $p$
\end{definition}

Let $\mathcal{L}$ be a light path tree of some light path $p$ in a maximum path decomposition $\mathcal{P}(T)$.
For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $T$
(i.e. $S_l(v)$ are the children of $v$ in $\mathcal{L}$).  Then define the cost of $v$ as 
$c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$.

We now define a weighted version of $\mathcal{L}$ based on these cost.  First we augment $\mathcal{L}$ by attaching one 
leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  Now for any vertex $u$ in this tree, define 
its weight to be the sum of the weights of its children (where at the leaves the weights are just the costs).  
To simplify notation, we will also refer to this new weighted tree as $\mathcal{L}$.

As the weight $w(v)$ includes the sizes of the heaps of all leaves in the subtree rooted at $v$, we have the following.

\begin{claim}
For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$
\end{claim}

Let $W= \sum_{v\in \mathcal{L}} w(v)$.  Suppose one could prove $W=O(H_r)$, where $r$ is the root of $\mathcal{L}$.  
Then by the above claim, 
$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, and so by convexity and \Cor{heavyLight}, we could argue the entire cost 
of all light paths in the light path tree can be charge to a heavy path.  So proving $W=O(H_r)$ is our new goal.

\pagebreak
\subsection{Junk, don't read}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{More Junk}

The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.
